Less Funding, More Lecturers, and Fewer Professors: \\ \vspace{0.1cm} Stagnating State Funding for Higher Education and its Effect on Faculty

### 2024-02-28 text/sections/03-concepts.tex
writing edits made.

Universities can respond multiple ways to the falls in per student state funding.
\cite{NBERw23736} established that public universities who experienced state funding cuts did not fully off-set by raising tuition, and cutting total spending.
But it is not clear how these spending cuts went on to affect faculty --- or whether they affected faculty at all.

Wider changes in the American higher education sector may concurrently explain the trends in state funding and faculty substitution.
College selectivity changed drastically over the 21st century, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, leaving public universities substituting toward lecturers over time.
This is one way in which the trends may be concurrent, but not causal.
I address this issue by using shift-share instrument approach to isolate funding cuts that affected more funding-intensive universities, measuring the rate of substitution between lecturers and professors in response to yearly changes in state-wide funding.

Universities could respond to funding cuts by cutting faculty salaries, leading to more faculty leaving for jobs at other universities.
Multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, and Cornell University implemented a professor salary cut in 2020 at the onset of the Covid-19 pandemic.\footnote{
    The Cornell University administration expected large fiscal squeezes in mid-2020, so imposed a faculty salary cut, and then returned the amount cut later in 2020 when the financial squeeze did not materialise.}
Funding cuts could also limit the salaries public universities can offer to new faculty hires, leading to fewer or less accomplished faculty accepting offers at public universities.\footnote{
    Faculty often consider their outside options in job decisions, considering offers for employment and/or promotion from multiple universities at the same time --- see \cite{blackaby2005} for an overview of faculty outside options.
    If public universities can only make low salary offers thanks to funding cuts, then they are less likely to land professors with multiple offers.
}
This effect may not be the same across faculty seniority; most universities hire more junior faculty each year, so that yearly hiring disruptions may disrupt the looser market for established full professors more than tighter market for assistant professors or lecturers.
While these factors make it unclear who will be affected most by funding cuts, there is one empirical fact worth noting: lecturers have substantially lower annual salaries than professors.
The average Illinois assistant professor's annual salary is $\$76,897$, tenured professors' is $\$109,283$, while lecturers' is $\$31,449$ (annual salary for 2010--2021 in 2021 USD, see \autoref{tab:illinois-summary}).
Secondly, hiring a new professor is a long-term costly investment, when compared to hiring a new lecturer.
Lecturers have shorter employment contracts than professors; assistant professors mostly work on 5--6 year long contracts, full professors on indefinite contracts (tenure), lecturers on yearly or termly contracts.
If a public university's primary obligation is to teach, and they must fulfil this objective with less funding, then substituting away from tenure-track and tenured professors towards lecturers is rationalisable purely on this basis.

It is not clear when the effects of state funding cuts will be realised.
Universities employ professors on multiple-year contracts (e.g., the tenure contract), and decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for effects to trickle down to faculty.
On the other hand, university departments hire in annual cycles, so that a funding cut in one year may immediately impact faculty hiring in the same year, and thus faculty counts in the next academic year.
Yet, it could be the case that faculty hiring resumes the very next year, or even hires more to compensate for last year's cancelled hiring cycle.
This would mean that state funding cuts leads to only transitory changes in faculty counts.
As such, this paper investigates the dynamics of state funding cuts and their effects on faculty, by employing local projections methods to measure the persistence of effects for multiple years following a state funding cut.

>>>>

Universities can respond multiple ways to state funding cuts.
\cite{NBERw23736} established that public universities experiencing state funding cuts did not fully off-set by raising tuition, and instead cut spending.
It is not clear how these funding or spending cuts went on to affect faculty --- or whether they affected faculty at all.

Wider changes in the American higher education sector could concurrently explain the trends in state funding and faculty substitution.
College selectivity changed drastically over the 21st century, where the top universities have become more selective while the average university less selective \citep{hoxby2009changing}.
While the average public university is becoming less selective, it is possible that their most productive or research-focused faculty more often move to more selective and prestigious (and often private) institutions, leaving public universities substituting toward lecturers over time.
This is one way in which the trends may be concurrent, but not causally related.
I address this issue by using shift-share IV approach to isolate funding cuts that affected more funding-intensive universities, measuring the rate of substitution between lecturers and professors in response to yearly changes in state-wide funding.

Universities could respond to funding cuts by cutting faculty salaries, leading to more faculty leaving for jobs at other universities.
Multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, and Cornell University implemented a professor salary cut in 2020 at the onset of the Covid-19 pandemic.\footnote{
    The Cornell University administration expected large fiscal squeezes in mid-2020, so imposed a faculty salary cut, and then returned the amount cut later in 2020 when the financial squeeze did not materialise.}
Funding cuts could also limit the salaries public universities can offer to new faculty hires, leading to fewer or less accomplished faculty accepting offers at public universities.\footnote{
    Faculty often consider their outside options in job decisions, considering offers for employment and/or promotion from multiple universities at the same time --- see \cite{blackaby2005} for an overview of faculty outside options.
    If public universities can only make low salary offers thanks to funding cuts, then they are less likely to land professors with multiple offers.
}
This effect may not be the same across faculty seniority; most universities hire more junior faculty each year, so that yearly hiring disruptions may disrupt the looser market for established full professors more than tighter market for assistant professors or lecturers.
While these factors make it unclear who will be affected most by funding cuts, there is one empirical fact worth noting: lecturers have substantially lower annual salaries than professors.
The average Illinois assistant professor's annual salary is $\$76,897$, tenured professors' is $\$109,283$, while lecturers' is $\$31,449$ (annual salary for 2010--2021 in 2021 USD, see \autoref{tab:illinois-summary}).
If a public university's primary obligation is to teach, and they must fulfil this objective with less funding, then substituting away from tenure-track and tenured professors towards lecturers is rationalisable purely on this basis.

If faculty are affected by state funding cuts, then it is still not clear whether effects are persistent or merely transitory.
Universities employ professors on multiple-year contracts (e.g., the tenure contract), and decisions to retire, move university, or leave academia are multiple-year commitments, so that it may take multiple years for effects to trickle down to faculty.
On the other hand, university departments hire in annual cycles, so that a funding cut in one year may immediately impact faculty hiring in the same year, and thus faculty counts in the next academic year.
% Yet, it could be the case that faculty hiring resumes the very next year, or even hires more to compensate for last year's cancelled hiring cycle.
%This means that state funding cuts could effect faculty in only transitory changes in faculty counts.
As such, this paper investigates the dynamics of state funding cuts and their effects on faculty, by employing local projections methods to measure the persistence of effects for multiple years following a state funding cut.

### 2024-02-28 text/sections/04-empirical.tex
writing edits made.

The system exploits the fact that institutions who rely on state funding more will be affected by shift-share instruments.
$Z_{i,t}$ is the instrument for state funding for institution $i$ in year $t$, interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per student) may arise as result of tuition hikes thanks declining state support, which would mean controlling for tuition would constitute a bad control.
    Nonetheless, estimates including tuition revenues (per student) as a control in the second stage of the instrumental variables estimates produces results of very similar magnitude and direction, and so are omitted.
}

The shift-share instrument is an instrument for state funding for each university in each year.  
The first-stage is then as follows, including institution and year fixed effects, where $X_{i,t}$ represents the amount of state funding divided by the number of full-time students attending the university.
\begin{equation}
    \label{eqn:firststage}
    X_{i,t} = \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t}
\end{equation}
The instrument must be conditionally independent of faculty outcomes for the instrumental variables results to be valid causal eestimates.
The instrument is conditionally independent if state policy decisions for funding of public universities are uncorrelated with unobserved institutional changes of any specific college or university in the state.
This assumption is plausible given that the majority of states have multiple (i.e., more than five) public universities, without any single university campus receiving the majority of state funding within any single state.
Secondly, the shift-share identification strategy requires exogeneity in either the base-line share or shift component of the instrument \citep{borusyak2022quasi}.
We satisfy the second: faculty or universitiy institutional-level decisions are not correlated with contemporaneous (or upcoming shocks) to state funding, so that the shift-share approach yields causal estimates by instrumental variables \citep{NBERw27885}.\footnote{
    It would be plausible to consider the case that universities make institutional-level decisions in a consistently different manner to those with differing reliance on state funding in 1990, so that exogeneity by the base-line share is not plausible here.
}
Additionally, the shift-share approach assumes that shift-share instruments affect faculty outcomes only via affecting university funding (exclusion restriction).

The conditional independence assumption for the shift-share instrument makes sense in theory for state funding, but is fundamentally untestable with data.
However, the assumption is bolstered by the fact that universities with higher values of the shift-share instrument differ little in observed characteristics compared to those with lower values \citep{pei2019poorly}.
Universities are by-in-large smaller (in terms of enrolment, total revenues, and professor count) at the top of the shift-share instrument distribution (see~\autoref{tab:summary-quantiles}).
Though on a per student basis, there is little difference across the distribution (except in the outcomes under consideration).
The shift-share instrument is positively associated with the total enrolment and total amount of state funding for each university, in both \$ and $\log$/percentage change terms, while other the other sources of university finances are not associated with the funding shock.
So that the other sources of finances are not clear confounders for the instrumental variables strategy, as they exhibit balance with respect to the funding shock instrument \citep{pei2019poorly}.
Yearly and individual fixed effects are included in regressions throughout to implicitly condition on mean differences between universities and years.

\subsection{Instrumental Variables Model}
\label{sec:iv-model-uni}

I use the instrument defined in \autoref{eqn:public-instrument} to overcome the endogeneity concerns for state funding to each public university, so the primary empirical model is an instrumental variables model.
\autoref{eqn:firststage} is the first stage for exogenous variation in the state funding for university $i$ in year $t$, and \autoref{eqn:secondstage} the second stage for the effect of state funding on faculty outcomes.\footnote{
    It is important to note the treatment effect isolated here; the instrumental variables approach identifies the local average treatment effect, one weighted to level of exposure when treatment is continuous as is the case here.
    So we interpret this treatment effect as a weighted average of effects on faculty at public universities to state funding changes, specific to shift-share instruments, among the complier group --- i.e., among universities who respond to funding shocks and would not have made faculty-outcome changes absent the funding shock.
    Also, we assume that no universities state funding increased in response to the negative shift-share instrument (monotonicity).
}
\begin{equation}
    \label{eqn:secondstage}
    Y_{i,t} = \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{equation}
I estimate the system by two stage least squares, including institution and year fixed effects.\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model implicitly accounts for yearly variation in professor count and university revenues arising from a university's size of growth/decline.
}
$Y_{i,t}$ represents faculty outcomes for university $i$ in year $t$, $\alpha_i, \gamma_t$ university and year fixed effects, and $\widehat X_{i,t}$ state funding for university $i$ estimated in first stage \eqref{eqn:firststage}.

@@

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per student, or in percent change rate?
The funding shock instrument is a strong predictor for the average university's level of state funding in either unit; a funding shock of \$1,000 per student in the entire state leads to around \$1,176 per student at the university, while a funding shock of $-10$\% leads to around 9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same to California and Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in their absolute funding for higher education in 1990, but most have experienced a decline of similar order to 30\%, so that the stagnation in state funding has been a percent change trend over this time period.
As such, I include regression specifications where the explanatory and outcome variables are $\log$ transformed, and refer to these when stating results in percent change terms.

\subsection{Effects in Years After the Initial Funding Shock}
\label{sec:local-projections}
The effects of a change in the universities funding may not be immediate, and faculty may be affected multiple years after a funding shock.
Yet, the funding shock instrument is significantly auto-correlated year-on-year; a large state-level funding shock in year $t$ is also likely to experience a large funding shock in year $t-1$.
Similarly, funding shocks to higher education in a year are highly correlated with state funding per student in the 5 years before and after the initial shock in IPEDS data (see \autoref{fig:lag-firststage}).
Thus, a linear model correlated state funding in year $t$ with outcomes in year $t+k$ will suffer from time-series confounding.

I employ a local projections approach to model how faculty outcomes in years $t+k$ are affected by state funding in year $t$, for future years $k > 0$ \citep{jorda2005}.
The local projections method is an empirical model used to estimate dynamic treatment effects when the treatment is not binary, so that time-series confounding (i.e., treatment in time $t$ is correlated with treatment in time $t-1$) is present \citep{montiel2021local}.
This method estimates the effect of state funding $X_{i,t}$ on faculty outcomes $Y_{i, t+ k}$ in follow-on years $k > 0$, while accounting for the auto-correlation between state funding in surrounding years.\footnote{
    The funding shock has a persistent effect on state funding, multiple years after the initial funding shock, so that the instrument is similarly strong for the local projections method (see \autoref{fig:firststage-lp}).
}
Additionally, the approach accommodates the shift-share instrument for state funding \citep{olea2021inference}, so that I present instrumental variables estimates for this method as in the rest of the empirical analysis.

>>>>

The system exploits the fact that institutions who rely more on state funding will be affected by state-wide funding cuts.
$Z_{i,t}$ is the shift-share instrument for institution $i$ in year $t$.
It is contructed by interacting the average funding for universities in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e., without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
$Z_{i,t}$ is negative to reflect the fact that the long term trend in, and most of the short-run shifts in, state funding for higher education have been negative.\footnote{
    When used in log terms, the instrument is the negative of the logged shock --- i.e., $- \log \left( -Z_{i,t} \right)$ as opposed to log of the negative shock $\log Z_{i,t}$ directly.
}
State funding has been falling, so that the instrument describes shocks to university revenues, mostly in a negative direction.\footnote{
    \label{foot:control}
    \cite{NBERw27885} note the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (not total revenue), as do \cite{NBERw27885}.
    On the other hand, rises in tuition revenues (per student) may arise as result of tuition hikes thanks declining state support, which would mean controlling for tuition would constitute a bad control.
    Nonetheless, estimates including tuition revenues (per student) as a control in the second stage of the IV estimates produce results of very similar magnitude and direction, and so are omitted.
}

The first-stage is then as follows, including institution and year fixed effects, where $X_{i,t}$ represents the amount of state funding divided by the number of full-time students attending the university.
\begin{equation}
    \label{eqn:firststage}
    X_{i,t} = \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t}
\end{equation}
The instrument must be conditionally independent of faculty outcomes for the IV results to be valid causal estimates.
The instrument is conditionally independent if state policy decisions for funding of public universities are uncorrelated with unobserved institutional changes of any specific college or university in the state.
This assumption is plausible given that the majority of states have multiple (i.e., more than five) public universities, without any single university campus receiving the majority of state funding within any single state.
Secondly, the shift-share identification strategy requires exogeneity in either the base-line share or shift component of the instrument \citep{borusyak2022quasi}.
We satisfy the second: faculty or university institutional-level decisions are not correlated with contemporaneous (or upcoming shocks) to state funding, so that the shift-share approach yields causal estimates by IV \citep{NBERw27885}.\footnote{
    It would be plausible to consider the case that universities make institutional-level decisions in a consistently different manner to those with differing reliance on state funding in 1990, so that exogeneity by the base-line share is not plausible here.
}
Additionally, the shift-share approach assumes that shift-share instruments affect faculty outcomes only via affecting university funding (exclusion restriction).
This means that state-wide changes in higher education funding are assumed to affect faculty at universities only by changing how much their university is funded.

The conditional independence assumption for the shift-share instrument makes sense in theory for state funding, but is fundamentally untestable with data.
However, the assumption is bolstered by the fact that universities with higher values of the shift-share instrument differ little in observed characteristics compared to those with lower values \citep{pei2019poorly}.
Universities are by-in-large smaller (in terms of enrolment, total revenues, and professor count) at the top of the shift-share instrument distribution (see~\autoref{tab:summary-quantiles}).
Though on a per student basis, there is little difference across the distribution (except in the outcomes under consideration).
The shift-share instrument is positively associated with the total enrolment and total amount of state funding for each university, in both \$ and $\log$/percentage change terms, while other the other sources of university finances are not associated with the funding shift-share.
So that the other sources of finances are not clear confounders for the IV strategy, as they exhibit balance with respect to the funding shift-share \citep{pei2019poorly}.
Yearly and individual fixed effects are included in regressions throughout to implicitly condition on mean differences between universities and years.

\subsection{Instrumental Variables (IV) Model}
\label{sec:iv-model-uni}

I use the instrument defined in \autoref{eqn:public-instrument} to overcome the endogeneity concerns for state funding to each public university, for an IV approach.
\autoref{eqn:firststage} is the first stage for exogenous variation in the state funding for university $i$ in year $t$, and \autoref{eqn:secondstage} the second stage for the effect of state funding on faculty outcomes.\footnote{
    The IV approach identifies the local average treatment effect, weighted towards larger values of the instrument in the continuous case.
    So we interpret this treatment effect as a weighted average of causal effects on faculty at public universities to state funding changes, specific to shift-share instruments, among compliers --- i.e., among universities who respond to funding cut and would not have made faculty-outcome changes absent the funding cut.
    Also, we assume that no universities state funding increased in response to the negative shift-share instrument (monotonicity).
}
\begin{equation}
    \label{eqn:secondstage}
    Y_{i,t} = \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{equation}
I estimate the system by two stage least squares, including institution and year fixed effects.\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model implicitly accounts for yearly variation in professor count and university revenues arising from a university's size of growth/decline.
}
$Y_{i,t}$ represents faculty outcomes for university $i$ in year $t$, $\alpha_i, \gamma_t$ university and year fixed effects, and $\widehat X_{i,t}$ state funding for university $i$ estimated in first stage \eqref{eqn:firststage}.

@@

It is not a priori clear which units are appropriate for this analysis;
does it make sense to consider state funding in purely dollar amounts per student, or in percent change rate?
The funding shift-share is a strong predictor for the average university's level of state funding in either unit; a funding shift-share of \$1,000 per student in the entire state leads to around \$1,176 per student at the university, while a funding shift-share of $-10$\% leads to around 9.77\% less state funding per year.
Yet, the level of state funding (and the outcome variables) vary greatly between states for the unit of analysis.
For example, the average Illinois public university receives \$10,709 in state funding per student in 1990 and  \$6,713 in 2021 (a fall of over 30\%), while California went from \$19,224 per student to \$12,915 in the same time span (a fall of 37\%).
While most states are not exactly the same as California or Illinois, this is example is telling for the phenomenon of stagnating state funding:
states vary in their absolute funding for higher education in 1990, but most have experienced a decline of around 30\% in per student funding, so that the trend is a common phenomenon in percentage terms across the US.
As such, I include regression specifications where the explanatory and outcome variables are $\log$ transformed, and refer to these when stating results in percent change terms.

\subsection{Effects in Years After the Initial Funding Cut}
\label{sec:local-projections}
The effects of a change in the universities funding may not be immediate, and faculty may be affected multiple years after a funding cut.
Yet, the funding shift-share instrument is significantly auto-correlated year-on-year; a large state-level funding cut in year $t$ is also likely to experience a large funding cut in year $t-1$.
\autoref{fig:lag-firststage} shows that the shift-share instrument is correlated with state funding five years into the past and future.
This level of time-series correlation would confound IV estimates of the effect of state funding in year $t$ on faculty outcomes in later years $t+k$, providing inconsistent estimates for the causal relationships.

I employ a local projections approach to model how faculty outcomes in years $t+k$ are affected by state funding in year $t$, for future years $k > 0$ \citep{jorda2005}.
The local projections method is an empirical model used to estimate dynamic treatment effects when the treatment is not binary, when time-series confounding is present \citep{montiel2021local}.
Faculty outcomes could be affected in only a transitory manner, where effects in later years are close to zero, or in a persistent manner, where the effects of state funding cuts linger for multiple years.
This method allows me to estimate the effect of state funding $X_{i,t}$ on faculty outcomes $Y_{i, t+ k}$ in follow-on years $k > 0$, credibly testing whether effects are transitory or long-lasting.
The approach accommodates the shift-share instrument for state funding \citep{olea2021inference}, so that I present results using the shift-share to instrument for state funding (as in the rest of the empirical analysis).\footnote{
    The funding shift-share has a persistent effect on state funding, multiple years after the initial funding shift-share, so that the instrument is similarly strong for the local projections method --- see \autoref{sec:appendix-ipeds-firststage} further evidence on the strength of the first stage in local projections estimation.
}
